---
title: "Resources"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, cache=FALSE)
knitr::opts_chunk$set(dev.args=list(bg="transparent"))    
knitr::opts_chunk$set(fig.retina = 2)
```

# Lectures

If you are interested in the lecture presentations as PDFs, you can download a Zip file containing all lectures here:
<a href="data_site/mareds_r_course_lectures.zip">Download lecture zip file [128.7 MB]</a>

# Dataset

The complete set of data files used in the lectures are available on the R Studio Server (for MARSYS students) but can also be downloaded here:
<a href="data_site/mareds_r_course_data.zip">Download data zip file [615 KB]</a>

<br>Alternatively, you can download the oceanographic and DATRAS fish data yourself from the [ICES Marine Data Portal](http://www.ices.dk/marine-data/data-portals).


# Packages

You should have the following packages installed on your computer to follow the exercises and demonstrations in the lectures (unless you will work on the institute's R Studio Server)

- [tidyverse](https://www.rdocumentation.org/packages/tidyverse/versions/1.2.1) (includes already: ggplot2, tibble, tidyr, readr, dplyr, stringr, forcats, purrr)
- [lubridate](https://www.rdocumentation.org/packages/lubridate/versions/1.7.0)
- [modelr](https://www.rdocumentation.org/packages/modelr/versions/0.1.1)
- [ggthemes](https://www.rdocumentation.org/packages/ggthemes/versions/3.4.0)
- [RColorBrewer](https://www.rdocumentation.org/packages/RColorBrewer/versions/1.1-2)
- [ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html)
- [ggmap](https://www.rdocumentation.org/packages/ggmap/versions/2.6.1)
- [grid](https://www.rdocumentation.org/packages/grid/versions/3.4.3)
- [gridExtra](https://www.rdocumentation.org/packages/gridExtra/versions/2.3)
- [cowplot](https://www.rdocumentation.org/packages/cowplot/versions/0.9.2)
- [FSA](https://www.rdocumentation.org/packages/FSA/versions/0.8.17)


All these packages are available on [CRAN](https://cran.r-project.org) and can be installed in one go using the `install.packages()` function:
```{r, eval=FALSE}
install.packages(c("tidyverse", "lubridate", "modelr", "ggthemes", "RColorBrewer", "ggmap", 
  "grid", "gridExtra", "cowplot", "FSA"))
```